A Cross-Domain Benchmark for Active Learning

### 2024-06-04 doc/al_benchmark_paper.tex
recommendations and pitfalls from baselines papers

@@

>>>>

Recommendation. Use the backbone architecture with
the community-accepted definition that is best suited
for the dataset at hand and consistently use it across all
experiments. In the image classification domain, we
suggest using ResNet18 for CIFAR-10 and CIFAR-100.

Recommendation. Control the type of optimizer across
methods for comparative evaluations to ensure that the
yield performance difference stems from an active learning method itself. As SGD often generalizes better, we
encourage its use for deep active learning.

Recommendation. Pragmatically fix the learning rate
to 0.1 for SGD on image datasets. While continuous
hyperparameter tuning can improve overall performance,
a fixed learning rate does not change the ranking of AL
methods from a comparative evaluation’s point of view.

Recommendation. One may use data augmentation if
applied consistently across methods, such that it does
not affect the overall ranking. However, a commonly
accepted baseline is needed, e.g., random horizontal flipping and random cropping for image classification.

Recommendation. Refine model parameters across AL
batch (“warm starts”) to prevent exhaustive reinitialization and feed initialization of the backbone model’s
weights and the “init sets” with fixed inputs over multiple runs to average out the randomness. Moreover, use
identical seeds for all methods under investigation.

Recommendation. Run experiments multiple times to
compensate for non-deterministic operations. If the resulting variance is larger than the gained improvement,
use deterministic operations stringently

Recommendation SW. Configure and verify influence parameter in active learning implementations thoroughly. To foster future research, we
provide implementations as part of our framework at:
https://intellisec.de/research/eval-al

Recommendation HW. Ensure that comparative evaluations are run on identical hardware. While it is not
necessary to execute all experiments on the same physical device, the GPU model, for instance, should be the
same. Do not mix hardware and list hardware details.

Recommendation. Consider multiple query-batch sizes
in the evaluation. The choice of the sizes needs to be
appropriate for the total number of unlabeled samples.

Recommendation. Compare active learning strategies
without sub-sampling, unless one of the approaches uses
it as a fundamental building block. In this case a detailed
analysis of the influence of sub-sampling is necessary.

Recommendation. Evaluate active learning strategies
on multiple benchmark datasets, that comprise balanced,
imbalanced, small-scale, and large-scale datasets to cover
most relevant cases in practice.

Recommendation. For a comprehensive analysis of
AL strategies, the overall comparative evaluation should
incorporate as many variables from Section 3 to yield a
summarized PPM that is as expressive as possible.

@@

P1 Data distribution. The proposed evaluation over a diverse selection of dataset distributions
including specific roll-out datasets proved essential for realistic evaluation of QMs as well as the
different training strategies. One main insight is the fact that class distribution is a crucial predictor
for the potential performance gains of AL on a dataset: Performance gains of AL are generally higher
on imbalanced datasets and occur consistently even for ST models with a small starting budget,
which are typically prone to experience cold start problems. This observation is consistent with a few
previous studies [30, 35, 54]. Further, our results underpin the importance of the roll-out datasets e.g.
when looking at the sub-random performance of BALD (with Self-SL) and Entropy (with ST) on
MIO-TCD. Such worst-case failures of AL application (increased compute and labeling effort due to
AL) could not have been predicted based on development data where all AL-parameters are optimized.
Another example is the lack of generalizability of Semi-SL, where performance in relation to other
Self-SL and ST decreases gradually with data complexity (going from CIFAR-10 to CIFAR-100/-10
LT to the two roll-out datasets MIO-TCD and ISIC-2019.)

P2 Starting budget. The comprehensive study of various starting budgets on all datasets reveals
that AL methods are more robust with regard to small starting budgets than previously reported
[6, 20, 43]. With the exception of Entropy we did not observe cold start problems even for any QM
even in combination with notoriously prone ST models. The described robustness is presumably
enabled by our thorough classifier configuration (P4) and heuristically adapted query sizes (P3). This
finding has great impact potential suggesting that AL can be applied at earlier points in the annotation
process thereby further reducing the labeling cost, especially in combination with BADGE which
performed consistently better or on par with Random. Similarly for Self-SL models, AL performs
well on small starting budgets with the exceptions of CIFAR-100 (BALD, Entropy) and MIO-TCD
(BALD, Core-Set, Entropy).

P3 Query size. Based on our evaluation of the query size we can empirically confirm its importance
with regard to 1) general AL performance and 2) counteracting the cold start problem. The are,
however, surprising findings indicating that the exact interaction between query size and performance
remains an open research question. For instance, we observe the cold start problem for BALD with
Self-SL on CIFAR-100 (∼ 50% accuracy at 2k labeled samples for query sizes of 500 (low-label
regime) and 1k (mid-label regime). On the other hand, in the high label regime (budget of 5k and
query size of 5k) ST and Self-SL models with similar accuracies of 50% and 60%, respectively,
benefit from BALD. Since cold start problems are commonly associated with large query sizes, this
finding seems counter-intuitive, but has been reported before although without further investigations
[6, 20, 43]. To gain a better understanding of this phenomenon and how QMs interact with query
size in the low-label regime, we performed a dedicated experiment series for Self-SL training on
CIFAR-100 and ISIC-2019 (see Figure 3 and Appendix F.2). Due to a significant improvement
of BALD on CIFAR-100 for even smaller query sizes from sub-Random to Random performance
and no worsening on ISIC-2019, we can conclude that small query sizes represent an effective
countermeasure against the cold-start problem for BALD contrary to the findings of [44, 58] and
currently not considered as a solution [6, 20, 43]. This potentially explains the gap between these
findings and more theoretical works which advertise using the smallest query size possible [19, 33].
Importantly, while smaller query sizes appear as a potential solution for the observed instabilities of
BALD, they seem to have no considerable effect on the other QMs. Moreover, our ablation reveals
that even in the low-label settings, BADGE is the overall most reliable of all compared QMs and
exhibits no sub-Random performance. This adds to the existing reports of the robustness of BADGE
for higher label regimes [4, 58].

P4 Classifer configuration. Our results show that method configuration on a properly sized validation
set is essential for realistic evaluation in AL. For instance, our method configuration has the same
effect on the classifier performance as increasing the number of labeled training samples by a factor
of ∼ 5 (e.g. our ST model reach approximately the same accuracy of ∼ 44% trained on 200 samples
compared to models in the AL study by Bengar et al. [6] trained on 1k samples). The effectiveness of
our proposed lightweight HP selection on the starting budget including only three parameters (Sec. 3)
is demonstrated by the fact that all our ST models substantially outperform respective models found
in relevant literature [6, 20, 30, 35, 43, 54, 55] where HP optimization is generally neglected. Details
for this comparison are provided in Appendix H. This raises the question of to which extent reported
AL advantages could have been achieved by simple classifier configurations. Further, our models
also generally outperform expensively configured models by Munjal et al. [44]. Thus, we conclude
that manually constraining the search space renders HP optimization feasible in practice without
decreasing performance and ensures performance gains by Active Learning are not overstated. The
importance of the proposed strategy to optimize HPs on the starting budget for each new dataset is
supported by the fact that the resulting configurations change across datasets.

P5 Alternative training paradigms. Based on our study benchmarking AL in the context of
both Self-SL and Semi-SL, we see that while Self-SL generally leads to improvements across all
experiments, Semi-SL only leads to considerably improved performance on the simpler datasets
CIFAR-10/100, on which Semi-SL methods are typically developed. Generally, models trained
with either of the two training paradigms receive a lower performance gain from AL (over random
querying) compared to ST. Crucially, Self-SL models converge around 2.5 times faster than ST
models, while the training time of Semi-SL models is around 80 times longer than ST and often yields
only small benefits over Self-SL or ST models. The fact that AL entails multiple training iterations
amplifies the computational burden of Semi-SL, rendering their combination prohibitively expensive
in most practical scenarios. Further, the fact that our Semi-SL models based on Fixmatch do not seem
to generalize to more complex datasets in our setting stands in stark contrast to conclusions drawn by
[20, 43] as to which the emergence of Semi-SL renders AL redundant. Interestingly, the exact settings
where Semi-SL does not provide benefits in our study are the ones where AL proved advantageous.
The described contradiction with the literature underlines the importance of our proposed protocol
testing for a method’s generalizability to unseen datasets. This is especially critical for Semi-SL,
which is known for unstable performance on noisy and class imbalanced datasets as noted by studies
focusing on Semi-SL [7, 22, 23, 29, 45, 56].

OPTIONAL - aus deren conclusion
Main empirical insights revealed by our study:
\begin{itemize}
\item Assessment of AL methods in the literature is substantially hampered by subpar classifier configurations, but meaningful assessment can be restored by our protocol for lightweight hyperparameter
tuning on the starting budget.
\item AL generally provides substantial gains in class-imbalanced settings.
\item BADGE is the best-performing QM across a realistic range of datasets, starting budgets, and query
sizes and exhibits no failure modes (i.e. sub-Random performance).
\item Combining AL with Self-SL considerably improves performance, shortens the training time, and
stabilizes optimization, especially in low-label settings.
\item FixMatch results indicate that Semi-SL methods perform well on datasets where they have been
developed, but may struggle to generalize to more realistic scenarios such as class imbalanced data.
Combining the paradigm with AL additionally suffers from extensive training times.
\item BALD with Self-SL pre-trained models benefits from smaller query sizes on small starting budgets
possibly circumventing the "cold start problem".
\end{itemize}

Take-aways for developing and proposing new Active Learning algorithms:
AL methods should be tested for generalizability on roll-out datasets and come with a clear recipe
for real-world application including how to adapt all design choices to new settings. Since the
expected benefit of AL on a new setting increases with lower application costs, we believe there is a
high potential for wide-spread real-world use of AL by reducing the two prevalent cost factors: 1)
Engineering costs, which are reduced by building easy-to-use AL tools. 2) Computational costs,
which are reduced by explicitly including methods that shorten the training time in AL.

Take-aways for the cost-benefit analysis of deploying Active Learning:
\begin{enumerate}
\item Since we identify BADGE as a robust query method exhibiting no sub-random-sampling performance across all tested settings, the potential harm of AL is minimized allowing to base decisions
around deploying AL on the described cost-benefit analysis.
\item The expected benefit is high in settings, where there is a mismatch between the task-specific
importance of individual classes and their frequency in the dataset (e.g. class-imbalanced datasets
in combination with tasks requiring a balanced classifier).
\item The expected benefit further increases with the labeling cost that will be reduced by AL. AL is thus
likely to yield a net benefit in settings with high label cost and low computational and engineering
costs.
\end{enumerate}
